Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-16 grove-paper.tex
finish merging last changes

Finally, holes are represented by the absence of a child.
For example, in \autoref{fig:A simple tree with a hole-b} the absence of a right edge coming from \vSimpleTimes{}
corresponds to the hole on the right of the~\texttt{*} in~\texttt{x * \hole}.

@@

To start our examples, Alice moves her cursor to the hole in~\texttt{x * \hole} in \autoref{fig:A simple tree with a hole-b}
and constructs the variable~\texttt{y} as shown in \autoref{fig:Construction-b}.
This action corresponds to creating \eConstructionY{} from \vSimpleTimes{} at the \texttt{R} position to a newly
created vertex, \vConstructionY{}, containing the variable reference~\texttt{y}.
The resulting graph, shown in \autoref{fig:Construction-b}, represents the expression \texttt{x * y}.

@@

Next Alice deletes~\texttt{x} so that the code becomes~\texttt{\hole{} * y}.
This is modeled by deleting \eSimpleX{} as shown in \autoref{fig:Change-a}.
Notice that \vSimpleX{} continues to exist, and if it had any children, those children would remain connected to it.
In our system, once a vertex is created it is never deleted.
This allows further manipulations of those vertices by other users as shown later in this section.
Note that we omit such orphaned vertexes from the remaining diagrams if they are not relevant to the exposition.

Once an edge with a particular identifier is deleted, it cannot be recreated.

Thus if Alice performed an ``undo'' on this deletion, Grove would create a fresh edge between \vSimpleTimes{} and \vSimpleX{}.

@@

\newcommand{\eWrapTimesWrapC}{\eWrapTimes}

@@

For example, in \autoref{fig:Move} Alice continues by
repositioning \texttt{\hole{} * y} from the left child of \texttt{+} to its right child.
This corresponds to the following sequence of graph edits, shown in \autoref{fig:Move}:
(a) delete \eWrapTimesWrapC{} leaving \vSimpleTimes{} temporarily orphaned and
(b) add \eMoveTimes{} to \vSimpleTimes{} in the right child position of \vWrapPlus{}.

@@

To start, Alice and Bob both have the graph in \autoref{fig:Move-b}, which
represents the code~\texttt{\hole{} + (\hole{} * y)}.

Alice adds~\vDifferentAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vConstructionY{} to \vDifferentBob{}.
Before transmitting their edits to each other,
Alice and Bob have the graphs in \autoref{fig:Editing Different Parts of the Code-a}
and \autoref{fig:Editing Different Parts of the Code-b}, respectively.
(Note that the transition edge from \autoref{fig:Move-b} to
\autoref{fig:Editing Different Parts of the Code-b} represents
multiple edges, i.e., deleting~\eConstructionY{} and adding~\eDifferentBob{} along with its child~\vDifferentBob{}.
We thus mark it with a star.)\todo{star size}\todo{"sync"}

@@

After Alice and Bob share the edits they made in the previous subsection,
Alice changes~\vDifferentAlice{} to~\vNestedAlice{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the right child of~\vWrapPlus{} to its left child
by deleting~\eMoveTimes{} and adding~\eNestedBob{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since edits are based on edge identities,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

@@

Next we address what happens when edits conflict.
For example, Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in Alice and Bob having the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
These graphs have two different edges for the~\texttt{R} position of~\vWrapPlus{}.
\autoref{fig:Multi-child-a} has~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and \autoref{fig:Multi-child-b} has~\eMultiChildBob{}, which points to~\vMultiChildBob{}.
We call this a \emph{multi-child conflict}.

@@

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w * \hole{}) + \hole{}}.
Alice moves~\vNestedAlice{} to the~\texttt{R} position of~\vSimpleTimes{},
while Bob moves it to the~\texttt{R} position of~\vWrapPlus{}.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} * w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} * \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedAlice{} and adds~\eMultiParentBob{}.
As explained in \autoref{sec:Formalism}, deletes are idempotent so Alice and Bob both deleting~\eNestedAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:Multi-parent-d}, in which~\vNestedAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole{} * \hole{}) * (\hole{} + \hole{})) + \hole}.
Alice moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole{} * \hole{}) + ((\hole{} + \hole{}) * \hole{})}.
On the other hand, Bob does the same
but puts~\vMultiCycleTimes{} under~\vMultiCyclePlus{}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole{} * \hole{}) + ((\hole{} * \hole{}) + \hole{})}.

@@

For example, consider the situation in \autoref{fig:Disconnection}.
\autoref{fig:Disconnection-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(\hole{} * \hole{}) + (\hole{}  * \hole{})}.
Alice moves~\vMultiCycleTimes{} under~\vSimpleTimes{}.
This results in \autoref{fig:Disconnection-b},
which represents the expression~\texttt{(\hole{} * (\hole{} * \hole{})) + \hole}.
On the other hand, Bob puts~\vSimpleTimes{} under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnection-c},
which represents the expression~\texttt{((\hole{} * \hole{}) * \hole{}) + \hole{}}.

@@

% XXX deletes are idempotent

>>>>

Finally, holes are represented by the absence of a child.
For example, in \autoref{fig:A simple tree with a hole-b} the absence of a \texttt{right} edge coming from \vSimpleTimes{}
corresponds to the hole on the right of the~\texttt{*} in~\texttt{x * \hole}.

@@

To start our examples, Alice moves her cursor to the hole in~\texttt{x * \hole} in \autoref{fig:A simple tree with a hole-b}
and constructs the variable~\texttt{y} as shown in \autoref{fig:Construction-b}.
This action corresponds to creating \eConstructionY{} from \vSimpleX{} at the \texttt{R} position to a newly
created vertex, \vConstructionY{}, containing the variable reference~\texttt{y}.
The resulting graph, shown in \autoref{fig:Construction-b}, represents the expression \texttt{x * y}.

@@

Next Alice deletes~\texttt{x} so that the code becomes~\texttt{\hole{} * y}.
This is modeled by deleting \eConstructionY{} as shown in \autoref{fig:Change-a}.
Notice that \vConstructionY{} continues to exist, and if it had any children, those children would remain connected to it.
In our system, once a vertex is created it is never deleted.
This allows further manipulations of those vertices by other users as shown later in this section.
Note that we omit such orphaned vertexes from the remaining diagrams if they are not relevant to the exposition.

Once an edge with a particular identifier is deleted, it cannot be recreated.
Thus if Alice performed an ``undo'' on this deletion, Grove would create a fresh edge between \vSimpleTimes{} and \vSimpleX{}.

@@

@@

For example, in \autoref{fig:Move} Alice continues by
repositioning \texttt{\hole{} * y} from the left child of \texttt{+} to its right child.
This corresponds to the following sequence of graph edits, shown in \autoref{fig:Move}:
(a) delete \eWrapPlus{} leaving \vSimpleTimes{} temporarily orphaned and
(b) add \eMoveTimes{} to \vSimpleTimes{} in the right child position of \vWrapPlus{}.

@@

To start, Alice and Bob both have the graph in \autoref{fig:Move-b}, which
represents the code~\texttt{\hole{} + (\hole{} + y)}.

Alice adds~\vDifferentAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vConstructionY{} to \vDifferentBob{}.
Before transmitting their edits to each other,
Alice and Bob have the graphs in \autoref{fig:Editing Different Parts of the Code-a}
and \autoref{fig:Editing Different Parts of the Code-b}, respectively.
(Note that thee transition edge from \autoref{fig:Move-b} to
\autoref{fig:Editing Different Parts of the Code-b} represents
multiple edges (i.e., deleting~\eConstructionY{} and adding~\eDifferentBob{} along with its child~\vDifferentBob{}).
We thus mark it with a star.)\todo{star size}\todo{"sync"}

@@

After Alice and Bob share the edits they made in the previous subsection,
Alice changes~\vDifferentAlice{} to~\vNestedAlice{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the left child of~\vWrapPlus{} to its right child
by deleting~\eMoveTimes{} and adding~\eNestedBob{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since edits are based on edge identities,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

@@

Next we address what happens whn edits conflict.
For example, Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in Alice and Bob having the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
These graphs have two different edges for the~\texttt{R} position of~\vWrapPlus{}.
\autoref{fig:Multi-child-a} has~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and \autoref{fig:Multi-child-b} has~\eMultiChildBob{}, which points to~\vMultiChildBob{}.
We call this a \emph{multi-child conflict}.

@@

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w + \hole{}) + \hole{}}.
Alice moves~\vNestedAlice{} to the~\texttt{R} position of~\vSimpleTimes{},
while Bob moves it to the~\texttt{R} position of~\vWrapPlus{}.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} + w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} + \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedAlice{} and adds~\eMultiParentBob{}.
As explained in \autoref{sec:Formalism}, deletes are idempotent so Alice and Bob both deleting~\eNestedAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:Multi-parent-d}, in which~\vNestedAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole{} + \hole{}) + (\hole{} * \hole{})) + \hole}.
Alice moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole{} + \hole{}) + ((\hole{} * \hole{}) + \hole{})}.
On the other hand, Bob does the same
but puts~\vMultiCycleTimes{} and under~\vMultiCyclePlus{}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole{} + \hole{}) + ((\hole{} + \hole{}) * \hole{})}.

@@

For example, consider the situation in \autoref{fig:Disconnection}.
\autoref{fig:Disconnection-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(\hole{} * \hole{}) + (\hole{}  * \hole{})}.
Alice moves~\vMultiCycleTimes{} under~\vSimpleTimes{}.
This results in \autoref{fig:Disconnection-b},
which represents the expression~\texttt{(\hole{} * (\hole{} * \hole{})) + \hole}.
On the other hand, Bob puts~\vSimpleTimes{} and under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnection-c},
which represents the expression~\texttt{\hole{} + ((\hole{} * \hole{}) * \hole{})}.

@@
